Hydrodynamics and rheology of adhesive vesicle doublets

### 2019-02-07 vesicle_code/docs/adhesion/adhesion.tex
Edited up to end of section II A

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Governing Equations \label{sec:ge}}
We consider suspensions of locally inextensible vesicles in an unbounded
two-dimensional viscous fluid.  For simplicity, we assume the fluid
viscosity both inside and outside the vesicles is constant, but our
method can be easily adjusted to account for a viscosity contrast.
Individual vesicles are denoted as $\gamma_j$, and they are
parameterized in arclength as $\xx_j(s,t)$.  The union of all vesicles
is denoted by $\gamma$.  Given a background velocity $\uu_\infty$, the
governing equations are
\begin{equation*}
\begin{aligned}
  \mu \Delta \uu = \nabla p, \quad &\xx \in \RR^2,
    &&\mbox{\em conservation of momentum} \\
  \nabla \cdot \uu = 0, \quad &\xx \in \RR^2, 
    &&\mbox{\em conservation of mass} \\
  \uu \rightarrow \uu_\infty, \quad &\|\xx\| \rightarrow \infty,
    &&\mbox{\em far-field condition} \\
  \uu(\xx,t) = \dot{\xx}, \quad &\xx \in \gamma,
    &&\mbox{\em velocity continuity} \\
  \xx_s \cdot \uu_s =0, \quad &\xx \in \gamma,
    &&\mbox{\em local inextensibility} \\
  \llbracket T \rrbracket \nn = \xxi, \quad &\xx \in \gamma,
    &&\mbox{\em non-zero traction jump}
\end{aligned}
\end{equation*}
where $\xxi$ is the traction jump that is the sum of the bending,
stretching, and adhesion forces.  The bending term $\BB \xx = -\kappa_b
\xx_{ssss}$ comes from the Helfrich energy, where $\kappa_b$ is the
bending rigidity modulus and $s$ is arclength.  For the numerical
examples, we set $\kappa_b = 1$ and consider adhesion strengths ranging
from $0.1$ to $1$ times the bending rigidity modulus.  The stretching
term is $\TT \sigma  = (\sigma \xx_s)_s$, where $\sigma$ is the tension,
and the tension acts as a Lagrange multiplier to satisfy the local
extensibility constraint.  The resistance to bending and stretching are
standard assumptions of vesicles.  In this work, we include an adhesive
term, $\AA \xx$, that we now describe.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Adhesion Model}
Sukumaran and Seifert~\cite{suk-sei2001} considers
adhesion between a single three-dimensional vesicle and a solid wall at
$z=0$.  They use the adhesion potential of the function form
\begin{align}
\label{eq:adhesion_potential}
  \phi(z) = H \left[ 
    \left(\frac{\delta}{z}\right)^m - \frac{m}{n} \left(\frac{\delta}{z}\right)^n \right],
\end{align}
where $z$ is the distance between the solid wall and the vesicle,
$\mathcal{H}$ is the Hamaker constant, $\delta$ is the adhesion length
scale and $(m,n) = (4,2)$.  We generalize this model to an unbounded
suspension of vesicles.  We assume that the adhesion force applies
between all pairs of points on different vesicles, so we are interested
in Hamaker constants that are appropriate for lipid-lipid interactions.
In Appendix~\ref{sec:appendixA}, the adhesive force at a point $\xx$ on
vesicle $j$ is shown to be
\begin{align}
  \AA\xx:=-\mathcal{H} m \delta^{n}\sum_{\substack{k=1 \\ k \neq j}}^M 
  \int_{\gamma_k} \frac{\xx - \yy}{\|\xx - \yy\|^{m+2}} 
  \left(\delta^{m-n} - \|\xx - \yy\|^{m-n} \right) ds_\yy.
  \label{eqn:adhesionForce}
\end{align}
The adhesion model is illustrated in Figure~\ref{fig:adhesionModel}.

@@

The exponents $(m,n)$ depend on the geometry of the two objects under
adhesion~\cite{Book_IntermolecularSurfaceForces}: $(m,n)=(4,2)$
corresponds to two flat, planar surfaces interacting with each under a
long-range attraction and a short-range repulsion. On the other hand,
$(m,n) = (12,6)$ corresponds to the L.-J. potential between two molecules.
In our case, the adhesion potential in the integral of
equation~\eqref{eqn:adhesionForce} is between two small patches of lipid
bilayer membrane (because the lipid molecules are coarse-grained in the
continuum modeling).  Thus a reasonable choice for $(m,n)$ between two
coarse-grained membrane patches would be between those for two planar
surfaces and two point molecules.

Large value of $m$ corresponds to a sharp increase in the repulsion
force as two objects are within the separation distance $\delta$.  This
poses great challenges for simulations as the problem becomes stiff
(i.e., requires a very small time step) for large $m$.  The adaptive
time-stepping BIE scheme makes it possible to simulate vesicle adhesive
dynamics with specified numerical precision for reasonable computation
time.  We explore several combinations of $(m,n)$ in the simulations of
two vesicles forming a doublet in a quiescent flow.  We found very
little difference in both the dynamic evolution and equilibrium
configuration between $(m,n)=(8,6)$ and $(m,n)=(4,2)$.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Governing Equations \label{sec:ge}}
We consider suspensions of locally inextensible vesicles in an unbounded
two-dimensional viscous fluid.  For simplicity, we assume the fluid
viscosity both inside and outside the vesicles is constant, but our
method can be easily adjusted to account for a viscosity contrast.
Individual vesicles are denoted as $\gamma_j$, $j=1,\ldots,M$, and they
are parameterized in arclength as $\xx_j(s,t)$.  The union of all
vesicles is denoted by $\gamma$.  Given a background velocity
$\uu_\infty$, the governing equations are
\begin{equation*}
\begin{aligned}
  \mu \Delta \uu = \nabla p, \quad &\xx \in \RR^2,
    &&\mbox{\em conservation of momentum} \\
  \nabla \cdot \uu = 0, \quad &\xx \in \RR^2, 
    &&\mbox{\em conservation of mass} \\
  \uu \rightarrow \uu_\infty, \quad &\|\xx\| \rightarrow \infty,
    &&\mbox{\em far-field condition} \\
  \uu(\xx,t) = \dot{\xx}, \quad &\xx \in \gamma,
    &&\mbox{\em velocity continuity} \\
  \xx_s \cdot \uu_s =0, \quad &\xx \in \gamma,
    &&\mbox{\em local inextensibility} \\
  \llbracket T \rrbracket \nn = \xxi, \quad &\xx \in \gamma,
    &&\mbox{\em non-zero traction jump}
\end{aligned}
\end{equation*}
where $\llbracket T \rrbracket$ is the jump in the stress tensor, and
$\xxi$ is the traction jump that is the sum of the bending, stretching,
and adhesion forces.  The bending force, arising from the Helfrich
energy model, is $\BB \xx = -\kappa_b \xx_{ssss}$, where $\kappa_b$ is
the bending rigidity modulus which we set to be 1 for all examples. The
stretching force is $\TT \sigma = (\sigma \xx_s)_s$, where the tension,
$\sigma$, acts as a Lagrange multiplier to satisfy the local
extensibility constraint.  The resistance to bending and stretching are
standard assumptions of vesicles.  In this work, we include an adhesive
force, $\AA \xx$, that we now describe.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Adhesion Model}
Sukumaran and Seifert~\cite{suk-sei2001} consider adhesion between a
single three-dimensional vesicle and a solid wall at $z=0$.  They use
the adhesion potential of the function form
\begin{align}
\label{eq:adhesion_potential}
  \phi(z) = H \left[ 
    \left(\frac{\delta}{z}\right)^m - \frac{m}{n} \left(\frac{\delta}{z}\right)^n \right],
\end{align}
where $z$ is the distance between the solid wall and the vesicle,
$\mathcal{H}$ is the Hamaker constant, $\delta$ is the adhesion length
scale, and $(m,n) = (4,2)$.  We generalize this model to an unbounded
suspension of vesicles. We assume that the adhesion force applies
between all pairs of points on different vesicles, so we are interested
in Hamaker constants that are appropriate for lipid-lipid interactions.
We use Hamaker constants ranging from $0.1$ to $1$ times the bending
rigidity modulus.  In Appendix~\ref{sec:appendixA}, the adhesive force
at a point $\xx$ on vesicle $j$ is shown to be
\begin{align}
  \AA\xx:=-\mathcal{H} m \delta^{n}\sum_{\substack{k=1 \\ k \neq j}}^M 
  \int_{\gamma_k} \frac{\xx - \yy}{\|\xx - \yy\|^{m+2}} 
  \left(\delta^{m-n} - \|\xx - \yy\|^{m-n} \right) ds_\yy.
  \label{eqn:adhesionForce}
\end{align}
The adhesion model is illustrated in Figure~\ref{fig:adhesionModel}.

@@

The exponents $(m,n)$ depend on the geometry of the two objects under
adhesion~\cite{Book_IntermolecularSurfaceForces}: $(m,n)=(4,2)$
corresponds to two flat, planar surfaces interacting with each under a
long-range attraction and a short-range repulsion. On the other hand,
$(m,n) = (12,6)$ corresponds to the L.-J.~potential between two molecules.
In our case, the adhesion potential in the integral of
equation~\eqref{eqn:adhesionForce} is between two small patches of lipid
bilayer membrane (because the lipid molecules are coarse-grained in the
continuum modeling).  Thus a reasonable choice for $(m,n)$ between two
coarse-grained membrane patches would be between those for two planar
surfaces and two point molecules.

Large value of $m$ corresponds to a sharp increase in the repulsion
force as two objects are within the separation distance $\delta$.  This
poses a numerical challenge since the problem becomes stiff (i.e.,
requires a very small time step) for large $m$.  The adaptive
time-stepping BIE scheme makes it possible to simulate vesicle adhesive
dynamics with specified numerical precision for reasonable computation
time.  We explore several combinations of $(m,n)$ in the simulations of
two vesicles forming a doublet in a quiescent flow.  We found very
little difference in both the dynamic evolution and equilibrium
configuration between $(m,n)=(8,6)$ and $(m,n)=(4,2)$.  Therefore, in
this work, we use $(m,n) = (4,2)$ to regulate the stiffness introduced
by the adhesive force.
